---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
# Assuming csv file is in the current workspace directory
data <- read.csv("activity.csv",colClasses=c("numeric","character","numeric"))
data$date<-as.Date(data$date)
# Creating a time variable to represent the interval. The date on this variable will not be used.
int<-data$interval
data$time<-strptime(paste(floor(int/100),":",int-100*floor(int/100),sep=""),format="%H:%M")
summary(data)
```

## What is mean total number of steps taken per day?
```{r}
library(plyr)
library(ggplot2)
stepsPerDay <- ddply(data,.(date),summarize,totalSteps=sum(steps,na.rm=TRUE))
qplot(totalSteps,data=stepsPerDay,binwidth=1000,xlab="Total Steps in One Day",main="Distribution of Total Steps Per Day from 10/1/2012 to 11/30/2012")
meanSteps<-prettyNum(mean(stepsPerDay$totalSteps,na.rm=TRUE),big.mark=",",digits=0)
medianSteps<-prettyNum(median(stepsPerDay$totalSteps,na.rm=TRUE),big.mark=",")
meanSteps
medianSteps
```


####*The mean total steps per day is `r meanSteps` and the median total steps per day is `r medianSteps`.*



## What is the average daily activity pattern?
```{r}
stepsPerInterval <- ddply(data,.(time),summarize,avgSteps=mean(steps,na.rm=TRUE))
with(stepsPerInterval,plot(time,avgSteps,type="l",xlab="Start Time of 5 Minute Interval", ylab="Average Number of Steps",main="Average Number of Steps per 5 Minute Interval"))
maxInterval <-stepsPerInterval$time[stepsPerInterval$avgSteps==max(stepsPerInterval$avgSteps)]
start<-format(maxInterval,format="%H:%M")
finish<-format(maxInterval+5*60,format="%H:%M")
start
finish
```



####*The 5 minute interval from `r start` to `r finish` contains the maximum average number of steps.*

## Imputing missing values

```{r}
numInts<-prettyNum(length(data$date),big.mark=",")
numMissing<-prettyNum(sum(is.na(data$steps)),big.mark=",")
numInts
numMissing
```

####*Of the `r numInts` intervals in this data set, `r numMissing` have missing values.*


```{r}
# For missing number of steps, I'm imputting the median of that time interval from the other days
medianStepsPerInterval <- ddply(data,.(interval),summarize,medSteps=median(steps,na.rm=TRUE))
inputedData<-data
inputedData$steps[is.na(data$steps)]<-sapply(data$interval[is.na(data$steps)], function(x) {medianStepsPerInterval$medSteps[medianStepsPerInterval$interval==x]})
stepsPerDayInputed <- ddply(inputedData,.(date),summarize,totalSteps=sum(steps,na.rm=TRUE))
qplot(totalSteps,data=stepsPerDayInputed,binwidth=1000,xlab="Total Steps in One Day (with Imputed Data)",main="Distribution of Total Steps Per Day from 10/1/2012 to 11/30/2012")
meanTotal<-prettyNum(mean(stepsPerDayInputed$totalSteps,na.rm=TRUE),big.mark=",",digits=0)
medianTotal<-prettyNum(median(stepsPerDayInputed$totalSteps,na.rm=TRUE),big.mark=",")
meanTotal
medianTotal
```


####*After data is inputed, the mean total steps per day is `r meanTotal` and the median total steps per day is `r medianTotal`. This mean is higher than it was before values were inputed because when missing values are ignored, many steps are missed. For example, 8 days had all missing values, so these days had totals of 0 steps. These zero totals pulled the mean down. The median, however, did not change because the total inputed for days with missing values was still small relative to the typical totals. Thus, the middle ranked value of these totals did not move.*

## Are there differences in activity patterns between weekdays and weekends?

```{r}
daytype<-factor(rep(NA,length(data$date)),levels=c("weekend","weekday"))
daytype[weekdays(data$date) %in% c("Sunday","Saturday")]<-"weekend"
daytype[!weekdays(data$date) %in% c("Sunday","Saturday")]<-"weekday"
inputedData$daytype<-daytype
byDaytype <- ddply(inputedData,.(time,daytype),summarize,avgSteps=mean(steps))
library(lattice)
byDaytype$minsPastMidnight<-as.numeric(format(byDaytype$time, "%H"))*60 + as.numeric(format(byDaytype$time, "%M"))
xyplot(avgSteps ~ minsPastMidnight | daytype, data = byDaytype, layout = c(1, 2),type="l",xlab="Time Interval in # Minutes Past Midnight",ylab="Number of Steps",main="Average Number of Steps by Time Interval")
meanWeekend<-round(mean(byDaytype$avgSteps[byDaytype$daytype=="weekend"]),0)
meanWeekday<-round(mean(byDaytype$avgSteps[byDaytype$daytype=="weekday"]),0)
meanWeekend
meanWeekday
```

####*Activity starts earlier on weekdays, but goes later on weekends. The overall activity appears slightly higher on weekends: the overall mean is `r meanWeekend` steps in 5 minutes on weekends versus `r meanWeekday` steps on Weekdays.*